---
title: "LimROTS"
author: "Ali Youssef"
format: html
editor: visual
---

# LimROTS 

An Extension of the ROTS Method with LIMMA Integration.

# UPS1 Case 0

This case study dataset includes DIA proteomics data derived from a UPS1-spiked E. coli protein mixture, processed using Spectronaut. UPS1 was spiked at eight different concentrations: 0.1 fmol, 0.25 fmol, 1 fmol, 2.5 fmol, 5 fmol, 10 fmol, 25 fmol, and 50 fmol, with three replicate injections for each concentration. In this case study, concentrations ranging from 0.1 fmol to 2.5 fmol are categorized as Conc. 2, while concentrations from 5 fmol to 50 fmol are categorized as Conc. 1. The dataset is available in a SummarizedExperiment format, along with the necessary metadata to run the example. This example demonstrates how LimROTS can be applied in a straightforward scenario where two groups are compared without including covariates.

```{r}
library(LimROTS)
library(parallel)

data("UPS1.Case0") 
print(UPS1.Case0)

set.seed(1234)
meta.info <- "Conc."
B <- 1000
K <- dim(UPS1.Case0)[1]/4
num_cores <- 10
cluster <- makeCluster(num_cores)
group.name <- "Conc."


limrots.result <- LimROTS(data.exp = UPS1.Case0, B = B, K = K, meta.info = meta.info, cluster = cluster, group.name = group.name, trend = TRUE, robust = TRUE)

```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
