---
title: "LimROTS with DREAM: Mixed Model Analysis for Repeated Measures 
        and Blocking Factors"
bibliography: references.bib
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 7
        toc: true
        toc_float: true
        toc_depth: 3
        number_sections: true
vignette: >
    %\VignetteIndexEntry{LimROTS with DREAM: Mixed Model Analysis}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

When samples share a common source — such as being processed by the same
laboratory instrument, batch, or subject — the observations are no longer
fully independent. Standard fixed-effects models treat all grouping
variables as fixed, which can inflate false discovery rates when a
grouping variable is better understood as a random sample from a broader
population of possible values.

DREAM (Differential expression for repeated measures)
([Hoffman & Roussos, 2020](https://doi.org/10.1093/bioinformatics/btaa687))
addresses this by fitting a linear mixed model per feature, where one or
more variables are modelled as random effects. This preserves the
within-group correlation structure rather than averaging it away, leading
to better-calibrated p-values and improved power when the random grouping
accounts for a meaningful portion of variation.

LimROTS integrates DREAM through the `variancePartition` package.
When the `formula.str` argument contains a random-effects term — i.e.,
any term of the form `(1|variable)` — LimROTS automatically routes the
analysis through `variancePartition::dream()` instead of the standard
`limma` workflow. This applies consistently to the main model fit,
bootstrap resampling, and permutation steps, so reproducibility
optimization remains valid under the mixed-model framework.

> **Scope:** The DREAM path in LimROTS currently supports two-group
> comparisons only. For more than two groups, use the `correlation_block`
> argument with a fixed-effects formula instead.

# UPS1 Case Study with DREAM

## Dataset Description

We reuse the UPS1 DIA proteomics dataset introduced in the main LimROTS
vignette. Briefly, it contains 1,656 proteins measured across 48
samples: 24 processed with Spectronaut and 24 with ScaffoldDIA, at two
concentration levels (`Conc. 1` — high, `Conc. 2` — low).

Because the two software tools (`tool`) process the same underlying
samples, the tool variable introduces a natural blocking structure.
Instead of including `tool` as a fixed covariate, we model it here as a
random effect using the formula `~ 0 + Conc. + (1|tool)`. This lets the
mixed model absorb the tool-level variance without consuming a degree of
freedom as a fixed contrast term.

```{r, echo=FALSE, message=FALSE, results="hide"}
library(LimROTS, quietly = TRUE)
data("UPS1.Case4")
```

```{r echo=FALSE}
table(UPS1.Case4$tool, UPS1.Case4$Conc.)
```

The expected result is that only the 48 UPS1 human proteins are truly
differentially expressed between concentration levels; no *E. coli*
protein should show a genuine concentration effect.

## Loading the Data

```{r load-packages, message=FALSE}
library(LimROTS, quietly = TRUE)
library(BiocParallel, quietly = TRUE)
library(ggplot2, quietly = TRUE)
library(SummarizedExperiment, quietly = TRUE)
```

```{r load-dataset}
data("UPS1.Case4")
print(UPS1.Case4)
```

## Running LimROTS with DREAM

The only change relative to the standard workflow is the `formula.str`:
the `tool` covariate is moved inside a random-effects term `(1|tool)`.
LimROTS detects the `|` character and automatically uses `dream()`
throughout the analysis.

Set a seed before running to ensure reproducibility.

```{r set-seed}
set.seed(1234, kind = "default", sample.kind = "default")
```

> **Note:** The group column in `colData` must be a `factor` with levels
> defined in the intended contrast order. For example, `Conc. 1` (high)
> vs `Conc. 2` (low) reports the fold-change as high − low. Define this
> explicitly if needed:
>
> ``` r
> colData(UPS1.Case4)$Conc. <- factor(
>     colData(UPS1.Case4)$Conc.,
>     levels = c("1", "2")
> )
> ```

```{r run-limrots-dream}
# Define analysis parameters
meta.info   <- c("Conc.", "tool")
niter       <- 100   # bootstrap iterations (use >= 1000 for real analyses)
K           <- 100   # top list size for reproducibility optimization

# Random-effects formula: tool is modelled as a random blocking factor
formula.str <- "~ 0 + Conc. + (1|tool)"

# Run LimROTS — DREAM is activated automatically by the (1|tool) term
UPS1.Case4 <- LimROTS(
    x           = UPS1.Case4,
    niter       = niter,
    K           = K,
    meta.info   = meta.info,
    group       = "Conc.",
    formula.str = formula.str,
    trend       = TRUE,
    robust      = TRUE,
    permutating.group = FALSE
)
```

> **Note:** `niter = 100` and `K = 100` are used here to reduce vignette
> build time. For production analyses, use `niter = 1000` or higher.
> We also recommend at least 4 parallel workers.

To increase the number of workers, pass a `BPPARAM` argument:

```{r bpparam-example}
# Windows
# BPPARAM <- SnowParam(workers = 4)

# Linux / macOS
# BPPARAM <- MulticoreParam(workers = 4)

# Pass to LimROTS via: LimROTS(..., BPPARAM = BPPARAM)
```

The results are appended directly to the `rowData` and `metadata` slots
of the input `SummarizedExperiment` object.

## Volcano Plot

The volcano plot shows corrected log-fold changes versus
$-\log_{10}$(q-value). At a 5% q-value threshold, LimROTS with DREAM
recovers the true UPS1 human proteins while controlling false positives
among the *E. coli* proteins.

```{r volcano-plot}
# Extract results
result_df <- data.frame(
    rowData(UPS1.Case4),
    row.names = rownames(UPS1.Case4)
)

# Annotate proteins
result_df$label <- ifelse(
    grepl("HUMAN", result_df$GeneID),
    "UPS1 (true positive)",
    "E. coli (true negative)"
)

ggplot(result_df, aes(
    x     = corrected.logfc,
    y     = -log10(qvalue),
    color = label
)) +
    geom_point(alpha = 0.7, size = 1.2) +
    geom_hline(
        yintercept = -log10(0.05),
        linetype   = "dashed",
        color      = "steelblue"
    ) +
    scale_color_manual(values = c(
        "E. coli (true negative)" = "grey60",
        "UPS1 (true positive)"    = "firebrick"
    )) +
    labs(
        title = "LimROTS with DREAM \u2014 UPS1 Case Study",
        x     = "Corrected Log Fold Change",
        y     = expression(-log[10](q~value)),
        color = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "top")
```

## Quality Control

```{r quality-control, results="hide", message=FALSE, warning=FALSE}
plot(metadata(UPS1.Case4)[["q_values"]])
hist(metadata(UPS1.Case4)[["q_values"]])
print(summary(metadata(UPS1.Case4)[["q_values"]]))
```

```{r session-info}
sessionInfo()
```

# References
