---
title: "LimROTS: Survival Analysis with Cox and Competing Risks Models"
bibliography: references.bib
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 7
        toc: true
        toc_float: true
        toc_depth: 3
        number_sections: true
vignette: >
    %\VignetteIndexEntry{LimROTS: Survival Analysis with Cox and Competing Risks Models}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

LimROTS extends its reproducibility-optimized testing framework to
survival analysis via the `LimROTS_survival()` function. It fits a Cox
proportional hazards model @survival — or, optionally, a competing risks
model via the Fine–Gray subdistribution approach @cmprsk — per feature
across bootstrap resamples. The reproducibility-optimization step then
selects the parameters $\alpha_1$ and $\alpha_2$ that maximize the
overlap of top-ranked features over those resamples, yielding:

$$t_{\alpha(p)} = \frac{\beta_{(p)}}{\alpha_1 + \alpha_2 \times SE_{(p)}}$$

where $\beta_{(p)}$ is the Cox model coefficient (log hazard ratio) and
$SE_{(p)}$ is its standard error for feature $p$.

P-values are derived from permutation-based null distributions using the
`qvalue` package @qvalue. This vignette demonstrates both models using a
Crohn's disease gut microbiome dataset.

# Citation

If you use `LimROTS` in your research, please cite:

> Anwar, A. M., Jeba, A., Lahti, L., & Coffey, E. (2025). LimROTS: A
> Hybrid Method Integrating Empirical Bayes and
> Reproducibility-Optimized Statistics for Robust Differential
> Expression Analysis. *Bioinformatics*, btaf570.
> <https://doi.org/10.1093/bioinformatics/btaf570>

# Dataset Description

We use the `crohn_survival` dataset from the `miaTime` package @miaTime.
This is a simulated dataset based on a real Crohn's disease microbiome
study [@Gevers2014; @coda4microbiome]. It contains **150 individuals**
and **48 gut microbial taxa** stored as a `TreeSummarizedExperiment`
object.

The relevant `colData` columns are:

| Column       | Description                                                |
|-----------------|-------------------------------------------------------|
| `diagnosis`  | Group label (`"CD"` = Crohn's disease, `"Control"`)        |
| `event`      | Binary event indicator (1 = Crohn's disease, 0 = censored) |
| `event_time` | Time to event or censoring (unitless)                      |

The analysis goal is to identify taxa whose abundance is associated with
time to Crohn's disease diagnosis.

## Loading the Data

```{r load-packages, message=FALSE}
library(LimROTS, quietly = TRUE)
library(miaTime, quietly = TRUE)
library(TreeSummarizedExperiment, quietly = TRUE)
library(BiocParallel, quietly = TRUE)
library(SummarizedExperiment, quietly = TRUE)
library(ggplot2, quietly = TRUE)
```

```{r load-dataset}
data("crohn_survival")
print(crohn_survival)
```

`crohn_survival` is a `TreeSummarizedExperiment`; coerce it to a plain
`SummarizedExperiment` as required by `LimROTS_survival()`:

```{r convert-to-se}
crohn_survival <- as(crohn_survival, "SummarizedExperiment")
print(crohn_survival)
```

Inspect the survival outcome distribution:

```{r inspect-coldata}
table(colData(crohn_survival)$diagnosis,
      colData(crohn_survival)$event)
```

`LimROTS_survival()` expects log-transformed assay values. The
`crohn_survival` assay contains relative abundance data; we apply a
centered log-ratio (CLR) transformation after adding a small pseudocount
to handle zeros:

```{r clr-transform}
counts <- assay(crohn_survival)
# Add pseudocount, compute CLR per sample
pseudocount <- 0.5
ra <- sweep(counts + pseudocount, 2, colSums(counts + pseudocount), "/")
clr <- t(apply(log(ra), 1, function(x) x - mean(x)))
# Replace the default assay so LimROTS_survival() uses CLR values
assay(crohn_survival) <- clr
hist(clr)
```

# Cox Proportional Hazards Analysis

The standard Cox model tests each taxon for an association with
time-to-event independently of any other covariates.

## Running `LimROTS_survival`

```{r set-seed-cox}
set.seed(1234, kind = "default", sample.kind = "default")
```

```{r run-cox}
crohn_cox <- LimROTS_survival(
    x           = crohn_survival,
    niter       = 100,    # use >= 1000 for real analyses
    K           = 10,     # number of top taxa
    meta.info   = c("event_time", "event"),
    formula.str = "~ Surv(event_time, event)",
    competing_risks   = FALSE,
    verbose           = FALSE
)
```

> **Note:** `niter = 100` is used here to reduce build time. For
> production analyses, use `niter = 1000` or higher and increase the
> number of parallel workers via the `BPPARAM` argument:
>
> ``` r
> # Windows
> BPPARAM <- SnowParam(workers = 4)
> # Linux / macOS
> BPPARAM <- MulticoreParam(workers = 4)
> # Pass to LimROTS_survival(..., BPPARAM = BPPARAM)
> ```

## Results

The results are appended to the `rowData` and `metadata` slots of the
returned `SummarizedExperiment`.

```{r cox-results}
cox_res <- data.frame(
    rowData(crohn_cox),
    row.names = rownames(crohn_cox)
)
head(cox_res[order(cox_res$pvalue), ])
```

Volcano plot of log hazard ratios versus $-\log_{10}$(q-value):

```{r volcano-cox}
cox_res$significant <- cox_res$qvalue < 0.05

ggplot(cox_res, aes(
    x     = log(exp_coef),
    y     = -log10(qvalue),
    color = significant
)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_hline(
        yintercept = -log10(0.05),
        linetype   = "dashed",
        color      = "steelblue"
    ) +
    scale_color_manual(
        values = c("FALSE" = "grey60", "TRUE" = "firebrick"),
        labels = c("FALSE" = "q \u2265 0.05", "TRUE" = "q < 0.05")
    ) +
    labs(
        title = "LimROTS Survival \u2014 Cox Model",
        x     = "Log Hazard Ratio",
        y     = expression(-log[10](q~value)),
        color = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "top")
```

# Competing Risks Analysis

When a competing event (e.g., death from another cause, remission)
prevents the event of interest from occurring, a competing risks model
is more appropriate than the standard Cox model. `LimROTS_survival()`
supports the Fine–Gray subdistribution hazard model @cmprsk when
`competing_risks = TRUE`.

The event variable must be coded as:

-   **0** — censored
-   **1** — event of interest
-   **2** — competing event

## Preparing the Data

The `crohn_survival` dataset has a binary event column (0/1). For
illustration, we recode half of the censored observations as having
experienced a competing event:

```{r prepare-competing}
crohn_cr <- crohn_survival

event_vec <- colData(crohn_cr)$event
censored_idx <- which(event_vec == 0)

set.seed(42)
competing_idx <- sample(censored_idx, floor(length(censored_idx) / 2))
event_vec[competing_idx] <- 2L

colData(crohn_cr)$event_cr <- event_vec
table(colData(crohn_cr)$event_cr)
```

> **Note:** In a real study, the competing event column should reflect a
> genuine competing risk recorded in the original data (e.g., surgical
> intervention, withdrawal). The recoding above is purely for
> demonstration purposes.

## Running `LimROTS_survival`

```{r set-seed-cr}
set.seed(1234, kind = "default", sample.kind = "default")
```

```{r run-competing}
crohn_cr <- LimROTS_survival(
    x           = crohn_cr,
    niter       = 100,
    K           = 48,
    meta.info   = c("event_time", "event_cr"),
    formula.str = "~ Surv(event_time, event_cr)",
    competing_risks   = TRUE,
    verbose           = FALSE
)
```

## Results

```{r cr-results}
cr_res <- data.frame(
    rowData(crohn_cr),
    row.names = rownames(crohn_cr)
)
head(cr_res[order(cr_res$pvalue), ])
```

Volcano plot for the competing risks model:

```{r volcano-cr}
cr_res$significant <- cr_res$qvalue < 0.05

ggplot(cr_res, aes(
    x     = log(exp_coef),
    y     = -log10(qvalue),
    color = significant
)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_hline(
        yintercept = -log10(0.05),
        linetype   = "dashed",
        color      = "steelblue"
    ) +
    scale_color_manual(
        values = c("FALSE" = "grey60", "TRUE" = "firebrick"),
        labels = c("FALSE" = "q \u2265 0.05", "TRUE" = "q < 0.05")
    ) +
    labs(
        title = "LimROTS Survival \u2014 Competing Risks Model",
        x     = "Log Subdistribution Hazard Ratio",
        y     = expression(-log[10](q~value)),
        color = NULL
    ) +
    theme_bw(base_size = 12) +
    theme(legend.position = "top")
```

# Quality Control

```{r qc-cox, results="hide", message=FALSE, warning=FALSE}
plot(metadata(crohn_cox)[["q_values"]])
hist(metadata(crohn_cox)[["q_values"]])
print(summary(metadata(crohn_cox)[["q_values"]]))
```

```{r session-info}
sessionInfo()
```

# References
